---
title: "Exploring Age-Related Immune–Vascular Changes in the Brain Using Single-Cell RNA-seq"
---
## Introduction
The brain, once thought to be immune privileged, is now recognized as a dynamic immune environment with a diverse array of resident and infiltrating immune cells. Microglia, the brain’s resident macrophages, are central to maintaining homeostasis, while border-associated immune populations—including perivascular macrophages, meningeal dendritic cells, and infiltrating T cells—play critical roles in surveillance and neuroimmune interactions.

## Dataset Overview
For my final project, I explored single-cell RNA-seq data generated by our lab from young and aged mouse brains. The dataset includes ~20,000 cells per age group, derived from whole-brain single-cell suspensions sorted for live cells and sequenced using the 10x Genomics platform.After sequencing, the data underwent standard quality control (QC) and preprocessing using the Seurat package in R. This included filtering low-quality cells, normalizing gene expression, identifying highly variable genes, and clustering cell populations.This analysis focuses on endothelial cells,  macrophages , and T cells— some of the key populations at the brain’s immune–vascular interface. By comparing age-related changes in the expression of adhesion molecules (e.g., ICAM1, VCAM1) and inflammatory mediators (e.g., CXCL10), the project aims to uncover how aging alters immune signaling and vascular function.

```{r}
# Read data file
library(readxl)
DEG_data <- read_excel("DEG_All_Condition_All_CellType 1.xlsx", sheet = 1)
```

```{r, message = FALSE, warning = FALSE}
# Load necessary packages
library(tidyverse)
library(dplyr)
library(ggplot2)
library(tidyr)
library(forcats)
```

```{r}
#look at the data
head (DEG_data)
```

```{r}
# Count number of entries per cell type (Cluster column)
all_cell_counts <- DEG_data %>%
  group_by(Cluster) %>%
  summarise(Count = n())

# Overview bar plot
ggplot(all_cell_counts, aes(x = reorder(Cluster, -Count), y = Count, fill = Cluster)) +
  geom_bar(stat = "identity") +
  labs(title = "Overview of All Cell Types in DEG Dataset",
       x = "Cell Type", y = "Number of Entries") +
  theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")
```


```{r}
# Define cell types I want to keep
cell_types <- c("Endothelial_cells", "Macrophages", "T_memory_cells")

# Filter the DEG dataset to keep only these 3 cell types
filtered_deg <- DEG_data %>%
  filter(Cluster %in% cell_types)
  
# only young vs. aged
young_vs_aged <- filtered_deg %>%
  filter(ident.1 == "G2_old", ident.2 == "G1_young")
table(young_vs_aged$Cluster)
```


```{r}

# Filter for endothelial cells aged vs young
endo_deg <- DEG_data %>%
  filter(Cluster == "Endothelial_cells",
         ident.1 == "G2_old", ident.2 == "G1_young") %>%
  mutate(log10_pval = -log10(p_val_adj),
         regulation = case_when(
           p_val_adj < 0.05 & avg_log2FC > 1 ~ "Upregulated",
           p_val_adj < 0.05 & avg_log2FC < -1 ~ "Downregulated",
           TRUE ~ "Not significant"
         ))

# Plot
ggplot(endo_deg, aes(x = avg_log2FC, y = log10_pval)) +
  geom_point(aes(color = regulation), alpha = 0.7, size = 2) +
  scale_color_manual(values = c(
    "Upregulated" = "firebrick2",
    "Downregulated" = "dodgerblue3",
    "Not significant" = "grey70"
  )) +
  labs(title = "Volcano Plot: Endothelial Cells (Aged vs Young)",
       x = "log2 Fold Change",
       y = "-log10 Adjusted p-value",
       color = "Regulation") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")

```


```{r}
# Filter for macrophages aged vs young
macrophage_deg <- DEG_data %>%
  filter(Cluster == "Macrophages",
         ident.1 == "G2_old", ident.2 == "G1_young") %>%
  mutate(log10_pval = -log10(p_val_adj),
         regulation = case_when(
           p_val_adj < 0.05 & avg_log2FC > 1 ~ "Upregulated",
           p_val_adj < 0.05 & avg_log2FC < -1 ~ "Downregulated",
           TRUE ~ "Not significant"
         ))

# Plot
ggplot(macrophage_deg, aes(x = avg_log2FC, y = log10_pval)) +
  geom_point(aes(color = regulation), alpha = 0.7, size = 2) +
  scale_color_manual(values = c(
    "Upregulated" = "firebrick2",
    "Downregulated" = "dodgerblue3",
    "Not significant" = "grey70"
  )) +
  labs(title = "Volcano Plot: Macrophages (Aged vs Young)",
       x = "log2 Fold Change",
       y = "-log10 Adjusted p-value",
       color = "Regulation") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")

```

```{r}
# Define top 5 DEGs per cluster
top5_combined <- DEG_data %>%
  filter(Cluster %in% c("Endothelial_cells", "Macrophages", "T_memory_cells"),
         ident.1 == "G2_old", ident.2 == "G1_young",
         !is.na(avg_log2FC), !is.na(p_val_adj),
         p_val_adj < 0.05,
         abs(avg_log2FC) > 1,
         !grepl("^Gm|^Rik|^ENSMUSG", gene_symbol)) %>%
  mutate(score = abs(avg_log2FC) * -log10(p_val_adj),
         display_gene = ifelse(is.na(gene_symbol) | gene_symbol == "", gene, gene_symbol),
         direction = case_when(
           avg_log2FC > 1  ~ "Up (old vs young)",
           avg_log2FC < -1 ~ "Down (old vs young)"
         )) %>%
  group_by(Cluster, direction) %>%
  slice_max(order_by = score, n = 5, with_ties = FALSE) %>%
  ungroup()

# Plot
ggplot(top5_combined,
       aes(x = fct_reorder(display_gene, avg_log2FC),
           y = avg_log2FC,
           color = direction)) +
  geom_point(size = 3) +
  geom_segment(aes(xend = display_gene, y = 0, yend = avg_log2FC),
               size = 0.6) +
  facet_wrap(~ Cluster, scales = "free_y") +
  coord_flip() +
  scale_color_manual(values = c("Up (old vs young)" = "firebrick2",
                                "Down (old vs young)" = "dodgerblue3")) +
  labs(title = "Top 5 Up & Down DEGs per Cell Type",
       x = "Gene", y = "log2FC (old vs young)", color = "") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")
```


```{r}

# Step 1: Define inflammatory/immune gene panel
inflammatory_genes <- c(
  "Il1b", "Il6", "Tnf", "Ifng", 
  "Ccl2", "Ccl3", "Ccl4", "Cxcl9", "Cxcl10",
  "Cd74", "Cd80", "Cd86", "Cd274", "Pdcd1lg2",
  "H2-Ab1", "H2-Aa", "H2-Eb1", "Tap1", "Tapbp",
  "Icam1", "Vcam1", "Sele", "Sell", 
  "Lyz2", "Nos2", "Ptprc",
  "March1", "Marchf1", "Fam49b", "Qki", "Niban1"
)

# Step 2: Filter DEG data
inflamm_subset <- DEG_data %>%
  filter(
    Cluster %in% cell_types,
    gene_symbol %in% inflammatory_genes,
    ident.1 == "G2_old", ident.2 == "G1_young",
    !is.na(avg_log2FC)
  )

# Step 3: Plot heatmap
ggplot(inflamm_subset, aes(x = Cluster, y = gene_symbol, fill = avg_log2FC)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(
    low = "blue", mid = "white", high = "red",
    midpoint = 0,
    limits = c(-10, 10),
    oob = scales::squish,
    name = "log2FC"
  ) +
  labs(
    title = "Log2 Fold Change of Inflammatory Genes (Aged vs. Young)",
    x = "Cell Type", y = "Gene"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 10),
    panel.grid = element_blank()
  )
```

## Gene Function Summary

These genes were selected for their roles in inflammation, immune regulation, or endothelial interactions:
| **Category**                       | **Representative Genes**                                                   | **Function**                                                                 |
|-----------------------------------|-----------------------------------------------------------------------------|------------------------------------------------------------------------------|
| Cytokines & Chemokines            | Tnf, Il1b, Il6, Ifng, Ccl2, Ccl3, Ccl4, Cxcl9, Cxcl10                       | Mediate inflammation, immune cell recruitment, and activation               |
| Co-stimulation & Checkpoints      | Cd80, Cd86, Cd274 (PD-L1), Pdcd1lg2 (PD-L2)                                 | Regulate T cell activation and inhibition                                   |
| Antigen Processing & Presentation | H2-Aa, H2-Ab1, H2-Eb1, Cd74, Tap1, Tapbp                                    | Present peptides via MHC molecules to T cells                               |
| Adhesion Molecules                | Icam1, Vcam1, Sele, Sell                                                    | Facilitate immune cell–endothelial interactions and trafficking             |
| Innate Immunity & Markers         | Lyz2, Nos2, Ptprc (CD45)                                                    | Mediate microbial killing and immune cell signaling                         |
| Immune Regulation & Stress Genes  | March1, Marchf1, Fam49b, Qki, Niban1                                        | Modulate antigen presentation, T cell activity, and cellular stress response |


## Conclusion
This project explored age-related gene expression changes in rare immune–vascular populations using single-cell RNA-seq data. Due to the low abundance of these cells, only a limited number of significant DEGs were identified—particularly in T cells, where no clear trends emerged.

In endothelial cells, a few adhesion molecules (e.g., VCAM1, ICAM1) appeared upregulated with age, suggesting possible endothelial activation. In macrophages, the volcano plot revealed many significant changes, though not all were directly related to inflammation. A more focused analysis of inflammatory pathways is needed to interpret these results biologically.

Given the limitations of whole-brain data, future work should involve sorted immune cell populations to improve resolution and detect clearer age-related transcriptional signatures.
